### 2022-10-10 AISTATS2023/paper.tex
review on simulation experiments

For achieving a computationally-grounded solution we set a threshold
\(K_{max}\) on the maximum number of bins which also discretizes the
solution space. The width of the bin can take discrete values that are
multiple of the minimum step
\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
solution, we use two indexes. The index
\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
\(j\)-th multiple of the minimum step, i.e., 
\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
\(T(i,j)\) is the cost of setting \(z_i=x_j\):
\begin{equation}
  \label{eq:recursive_cost}
  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
\end{equation}
%
where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
with limits \([x_l, x_j)]\):

@@

Simulation examples provide the freedom to select the data-generating
distribution \(p(\mathbf{X})\) and the predictive function
\(f(\cdot)\). We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for empirically
evaluating UALE against a solid ground-truth. The first group of
synthetic examples (Section~\ref{sec:simulation-examples-1}) compares
UALE with PDP-ICE plots in quantifying the average and the uncertainy
of the effect. The second group
(Section~\ref{sec:simulation-examples-2}) evaluates the accuracy of
the approximation (average effect and uncertainty) using the automatic
bin-splitting algorithm. In both groups, we illustrate the most
indicative examples; a more extensive evaluation is provided in the
Appendix.

\subsection{Case 1: Uncertainty Quantification}
\label{sec:simulation-examples-1}

In this simulation, we will compare UALE method against PDP-ICE in
quantifying the main effect and the uncertainty. The samples that we
use in all examples are coming from the following distribution:
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We create the following
piecewise linear function:

% Since there is
% ambiguity about the ground-truth first-order effect in cases of
% correlated features,
% e.g.~\citep{apley2020visualizing,Gromping2020MAEP},

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and
\(f_{\mathtt{int}}\) interacts the two non-correlated features.  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
equal weights (\(a_1=a_2\)).

In all cases, we provide the ground-truth average effect and
uncertainty computed with analytical derivations (proofs at the
Appendix) and we compare it against the approximation provided by each
method. The approximation is computed after generating \(N=500\)
samples.

@@

We test the feature effect when no interaction exits \(\alpha=0\) and
the weights are equal \(a_1=a_2=1\). The ground truth effect
\(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) when \(0 \leq x_1 < 0.25\),
\(-x_1\) when \(0.25 \leq x_1 < 0.5\) and zero otherwise. There zero
heterogeneous effects, therefore
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, ALE captures correctly the average effect and the zero
uncertainty and it also groups perfectly the constant-effect regions.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
  \caption{No interaction, Equal weights: Feature effect for \(x_1\);
    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:synth-ex-1-case-1}
\end{figure}

\paragraph{No Interaction, Different weights.}

As before, no interaction exits \(\alpha=0\), therefore, the
ground-truth uncertainty is zero, i.e.
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). The weigts are
\(a_1=2, a_2=0.5\), therefore, the ground-truth effect is
\(f_\mu^{\mathtt{GT}}(x_1)\) is, \(2x_1\) when \(0 \leq x_1 < 0.2\),
\(-2x_1\) when \(0.2 \leq x_1 < 0.4\) and zero otherwise. In
Figure~\ref{fig:ex-synth-1-2}, we observe that PDP estimation is
completely opposite to the ground-truth effect, i.e.~negative in the
region \([0, 0.2)\) and positive in \([0.2, 0.4)\), and the ICE
erroneously implies the existence of heterogeneous effects. As before,
ALE quantifies correctly the ground truth effect, the
zero-uncertainty and extracts the constant effect regions.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
  \caption{No interaction, Different weights: Feature effect for \(x_1\);
    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:ex-synth-1-2}
\end{figure}

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set the weights to \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects for features \(x_1, x_3\), because the local
effects at \(x_1\) depend on the unknown value of \(X_3\) and
vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
not appear in any interaction term. As we observe in
Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
and the uncertainty in all cases. PDP-ICE quantifies correctly the
effect and the uncertainty only in the case of feature \(x_3\),
because \(X_3\) is independent from other features. For features
\(x_1, x_2\) that are correlated the average effect of PDP
and the uncertainty of ICE plots are wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
  \caption{With interaction, equal weights: Feature effect for all
    features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
    (a) UALE method at the left columns and (b) PDP-ICE at the right
    column.}
  \label{fig:ex-synth-1-3}
\end{figure}

@@

Despite the model's simplicity, PDP-ICE fail in quantifying the
average effect and the uncertainty due to correlation between features
\(X_1\) and \(X_2\). UALE quantifies them correctly and estimates them
accurately extracting the regions with constant effect. The examples
above do not cover the case of an interaction term between correlated
features, e.g. \(x_1x_2\), because there is an open debate about the
ground-truth effect in this case~\citep{Gromping2020MAEP}.

@@

In this simulation, we aim to to quantify the advantages of automatic
bin-splitting. We set-up the following validation framework. We
generate a big dataset with dense sampling (\(N=10^6\)) and we treat
the DALE estimation with dense fixed-size bins (\(K=10^3\)) as
ground-truth. Afterwards, we generate less samples (\(N=500\)) and we
compare the fixed-size DALE estimation (for many different \(K\))
versus the automatic bin-splitting. In all cases, we use a bivariate
black-box function \(f(\cdot)\) where the samples are instances of the
distribution \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
splitting with \(K\) bins.  For consistent results, in all the
examples below, we regenerate samples and repeat the computations for
\(t = 30\) times, providing the mean value for all metrics.

@@

The evaluation is done with regard to two metrics that count the mean
bin error, where bin error is the absolute difference of the
approximation from the ground-truth. The metric
\(\mathcal{L}_{\mathtt{DP|K}}^{\mu} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \mu_k - \hat{\mu}_k | \) counts the
mean bin error on the average effect and
\(\mathcal{L}_{\mathtt{DP|K}}^{\sigma} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \sigma_k - \hat{\sigma}_k | \) the
mean bin error on the uncertainty. In both cases the ground truth is
computed from the dense bins of ground-truth that overlapse with the
wider bin of the approximation. We also provide the mean (per bin)
error term
\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} \rho_k \) for a complete understanding
of the uncertainty error.

@@

In this example, we define \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) with
\(5\) piecewise-linear regions of different-size, i.e., \(a_1\) equals
to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined by the
sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
Figure~\ref{fig:ex-synth-2-1}, UALE method extracts a sequence of
intervals with better \(\mathcal{L}_{\mathtt{DP}}^{\mu}\) and
\(\mathcal{L}_{\mathtt{DP}}^{\sigma^2}\) error compared to any
fixed-size splitting. Analyzing the fixed-size errors helps us
understand the importance of variable-size splitting. In
Figure~\ref{fig:ex-synth-2-1}(b), we observe a positive trend between
\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and \(K\), concluding that bin
effect estimation is more incosistent as \(\Delta x\) becomes smaller,
due to less points contributing to each bin. The interpretation of
variance error is slightly more complex. Given that the smallest
interval is \(\Delta x = 0.05 \Rightarrow K = 20\) and all intervals
are multiples of the smallest interval, any \(K\) that is not a
multiple of \(20\) adds nuisance uncertainty
\(\mathcal{L}^{\rho^2}_{\mathtt{K}}\) leading to a high variance error
\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\). In these cases, the variance
error reduces as \(K\) grows bigger because the length of the bins
that lie in the limit between two piecewise linear regions becomes
smaller. For \(K=\{20, 40, 60, 80, 100\}\) where
\(\mathcal{L}^{\rho^2}_{\mathtt{K}} \approx 0\), we conclude the same
as with with the mean effect error, i.e. the estimation becomes more
incosistent as \(K\) grows larger.

Variable-size extracts correctly the fine-grain bins, e.g., intervals
\([0.4, 0.45]\), \([0.45, 0.5]\), and acts as a merging mechanism to
create wider bins when the effect remains constant, e.g. interval
\([0.5, 1]\), leading to an optimal solution.

@@

In this example, we define a black-box function
\(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), where the effect is
non-linear in all the range of \(x\). This case has two
specialties. First, there is no obvious advantage of variable-size
versus fixed-size splitting, and, second, there is not an apriori
optimal bin-size. Widening a bin will increase the resolution error
\(\mathcal{L}^{\rho^2}\) and narrowing will make less robust. In
Figure~\ref{fig:ex-synth-2-2}, we observe that automatic bin splitting
finds a solution that compromises the conflicting objectives, i.e., it
keeps as low as possible both the main effect
\(\mathcal{L}^{\mu}_{\mathtt{K}}\) and the variance error
\(\mathcal{L}^{\sigma^2}_{\mathtt{K}}\).

@@

We exclude instances with missing values or with outlier values. As
outlier we define the feature values which ar over three standard
deviations away from the mean feature value. We also normalize all
features to zero-mean and unit standard deviation. We split the
dataset into \(N_{tr} = 15639\) training and \(N_{test} = 3910\) test
examples (80/20 split) and we fit a Neural Network with 3 hidden
layers of 256, 128 and 36 units respectively. After 15 epochs using
the Adam optimizer with learning rate \(\eta = 0.02\), the model
achieves a normalized mean squared error (R-Squared) of \(0.25\)
(0.75), which corresponds to a MAE of \(37 \cdot 10^3\) dollars.

@@

add limitations.

>>>>

\todo{Review all the description}
For achieving a computationally-grounded solution we set a threshold
\(K_{max}\) on the maximum number of bins which also discretizes the
solution space. The width of the bin can take discrete values that are
multiple of the minimum step
\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
solution, we use two indexes. The index
\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
\(j\)-th multiple of the minimum step, i.e., 
\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
\(T(i,j)\) is the cost of setting \(z_i=x_j\):
\begin{equation}
  \label{eq:recursive_cost}
  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
\end{equation}
%
where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
with limits \([x_l, x_j)]\):

@@

In XAI it is common~\citep{aas2021explaining, herbinger2022repid} to
use simulation examples for compering methods, because the
data-generating distribution \(p(\mathbf{X})\) and the predictive
function \(f(\cdot)\) are known, and, therefore, the ground-truth is
accesible. We follow this common XAI practice and we split the
evaluation in two groups of examples. The first group
(Section~\ref{sec:simulation-examples-1}) compares UALE definition
against PDP-ICE in modeling the average effect and the
uncertainty. The second group
(Section~\ref{sec:simulation-examples-2}) compares UALE approximation
using the automatic bin-splitting against the fixed-size alternative,
in estimating the average effect and the uncertainty.

\subsection{Case 1: UALE vs PDP-ICE}
\label{sec:simulation-examples-1}

In this simulation, we compare UALE against PDP-ICE. The aim of this
example is to highlight that when features are correlated, PDP and ICE
fail in quantifying the average effect and the heterogeneity, even
when the black-box function is relatatively simple.

We use the following distribution:
\(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\) and the following piecewise
linear function:

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{\mathtt{int}}(\mathbf{x}) = x_1x_3\). The linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the
interaction term \(f_{\mathtt{int}}\) the two non-correlated.  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha \neq 0\)) with
equal weights (\(a_1=a_2\)). In all cases, we provide the ground-truth
average effect and uncertainty computed with analytical derivations
(proofs at the Appendix) and we compare it against the approximations
provided by each method, after generating \(N=500\) samples.

@@

In this set-up, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\)
(equal weights). The ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\)
is: \(x_1\) in \([0, 0.25]\), \(-x_1\) in \([0.25, 0.5]\) and \(0\) in
\([0.5, 1]\). The uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) in \([0,1]\) because \(x_1\)
does not interact with ohter variables. In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, UALE models correctly both the average effect and the zero
uncertainty and approximates it accurately finding a zero aggregate
residual error partitioning of \(x_1\) domain.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
  \caption{No interaction, Equal weights: Feature effect for \(x_1\)
    using UALE (Left) and PDP-ICE (Right).}
  \label{fig:synth-ex-1-case-1}
\end{figure}

\paragraph{No Interaction, Non-Equal Weights.}

In this set-up, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= 0.5\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, 0.2]\), \(-2x_1\) in \([0.2, 0.4]\) and \(0\) in \([0.4,
1]\). As before, the ground-truth uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with other features.  In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is completely opposite to the ground-truth
effect, i.e.~negative in the region \([0, 0.2)\), positive in
\([0.2, 0.4)\), and the ICE erroneously implies the existence of
heterogeneous effects. As before, ALE quantifies correctly the ground
truth effect, the zero-uncertainty and partioning correctly the
\(x_1\) domain.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
  \caption{No interaction, Different weights: Feature effect for \(x_1\)
    using UALE (Left) and PDP-ICE (Right).}
  \label{fig:ex-synth-1-2}
\end{figure}

@@

In this set-up, we activate the interaction term, i.e.  \(a=1\),
keeping the weights equal \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects for features \(x_1\) and \(x_3\), because the
local effects at \(x_1\) depend on the unknown value of \(X_3\) and
vice-versa. The effect of \(x_2\) has zero-uncertainty since it does
not appear in any interaction term.  As we observe in
Figure~\ref{fig:ex-synth-1-3}, UALE correcly models the average effect
and the uncertainty in all cases.  On the other hand, PDP-ICE
quantifies correctly the average effect and the uncertainty only in
the case of feature \(x_3\), because \(X_3\) is independent from other
features. For the correlated features (\(x_1, x_2\)) both the average
effect (PDP) and the uncertainty (ICE) are wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
  \caption{With interaction, equal weights: From top to bottom,
    feature effect for features \(\{x_1, x_2, x_3\}\) using UALE (left
    column) and PDP-ICE (right column).}
  \label{fig:ex-synth-1-3}
\end{figure}

@@

Despite the model's simplicity, PDP-ICE fail in modeling both the
average effect and the uncertainty. The error is due to the method's
formulation, i.e.~the use of the marginal distribution that ignores
correlations between features, and not due to poor approximation
because of limited samples. In contrast, UALE models both correctly
and estimates them accurately extracting the regions with constant
effect.  The examples above do not cover the case of an interaction
term between correlated features, for example a term \(x_1x_2\),
because there is an open debate about the ground-truth effect in this
case~\citep{Gromping2020MAEP}.

@@

In this simulation, we illustrate the advantages of automatic
bin-splitting against the fixed-size alternative. For this reason, we
generate a very big dataset applying dense sampling (\(N=10^6\)) and
we treat the estimation with dense fixed-size bins (\(K=10^3\)) as the
ground-truth UALE. Afterwards, we generate less samples (\(N=500\))
and we compare the fixed-size estimation (for many different \(K\))
against UALE automatic bin-splitting. In all set-ups, we sample from
\(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^{\mathtt{DP}}} = \{z^{\mathtt{DP}}_{k-1}, \cdots,
z^{\mathtt{DP}}_{K}\}\) the sequence obtained by automatic
bin-splitting and with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size
splitting with \(K\) bins. The evaluation metrics we report are the
average number of \(t = 30\) independent runs, using each time
\(N=500\) different samples.

@@

The evaluation is done with regard to two metrics counting the mean
error per bin, where the bin error is definde as the absolute
difference of the approximation from the ground-truth. The first
metric, Eq.~\eqref{eq:eval_met_1}, counts the mean bin error wrt
average effect and the second, Eq.~\eqref{eq:eval_met_2} the mean bin
error wrt uncertainty:

\begin{equation}
  \label{eq:eval_met_1}
  \mathcal{L}_{\mathtt{DP|K}}^{\mu} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} | \mu(z_k- z_{k-1}) - \hat{\mu}(z_k-
  z_{k-1}) |
\end{equation}

\begin{equation}
  \label{eq:eval_met_2}
  \mathcal{L}_{\mathtt{DP|K}}^{\sigma} = \frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
    \mathcal{Z}^{\mathtt{DP|K}}} | \sigma(z_k- z_{k-1}) - \hat{\sigma}(z_k-
  z_{k-1}) |
\end{equation}

The ground truth \(\mu_k\) (\(\sigma_k\)) is the mean value of the
bin-effects (bin-uncertainties) of the bins inside the interval
defined by the wide bin. For better interpretation of the uncertainty
error, we also provide the mean (per bin) error term
\(\mathcal{L}^{\rho}_{\mathtt{DP|K}} =
\frac{1}{|\mathcal{Z}^{\mathtt{DP|K}}|} \sum_{k \in
  \mathcal{Z}^{\mathtt{DP|K}}} \mathcal{\Epsilon}(z_{k-1}, z_k) \).

@@

In this set-up, \(f(\mathbf{x}) = a_1x_1 + x_1x_2\) is a
piecewise-linear function with \(5\) different-width regions, i.e.,
\(a_1\) equals to \(=\{2, -2, 5, -10, 0.5\}\) in the intervals defined
by the sequence \(\{0, 0.2, 0.4, 0.45, 0.5, 1\}\). As we observe, in
Figure~\ref{fig:ex-synth-2-1}, UALE's approximation is better than all
fixed-size alternatives, in both mean effect (\(\mathcal{L}^{\mu}\))
and uncertainty (\(\mathcal{L}^{\sigma}\)). For understanding the
importance of variable-size splitting, we analyze the fixed-bin
error. In the top-right of Figure~\ref{fig:ex-synth-2-1}, we observe
the Law of Large Numbers (LLN); \(\mathcal{L}^{\mu}_{\mathtt{K}}\) and
\(K\) have positive correlation because as \(\Delta x\) becomes
smaller less points contributing to each estimation. The
interpretation of \(\mathcal{L}^{\sigma}\) is more complex. Given that
the smallest piecewise-linear region is \(\Delta x = 0.05 \) any
\(K\) that is not a multiple of \(\frac{1}{\Delta x} = 20\), adds
nuisance uncertainty \(\mathcal{L}^{\rho}_{\mathtt{K}} > 0\) leading
to a biased uncertainty approximation. For these \(K\), uncertainy
error decreases for bigger \(K\) because the length of the bins that
lie in the limit between two piecewise linear regions becomes
smaller. For \(K\) that are multiple of \(20\), i.e.
\(K=\{20, 40, 60, 80, 100\}\), the uncertainty approximation is
unbiased \(\mathcal{L}^{\rho}_{\mathtt{K}} \approx 0\). Here, as in
\(\mathcal{L}^{\mu}\), uncertainty estimation is worse as \(K\) grows
larger (LLN). As we observe in the top left of
Figure~\ref{fig:ex-synth-2-1}, automatic bin-splitting separates the
fine-grain bins, e.g.~regions \([0.4, 0.45]\), \([0.45, 0.5]\), and
unites constant-effect regions into a single bin, e.g.~region
\([0.5, 1]\).

@@

In this set-up, \(f(\mathbf{x}) = 4x_1^2 + x_2^2 + x_1x_2\), so the
effect is non-linear in \(0 \leq x_1 \leq 1\). In this set-up, wide
bins increase \(\mathcal{L}^{\rho^2}\) making the uncertainty
approximation more biased and narrow bins lead to a worse
approximation. In Figure~\ref{fig:ex-synth-2-2}, we observe that
automatic bin splitting manages to compromise the
conflicting objectives, \todo{Explain better}

@@

We exclude instances with missing and outlier values. As outlier we
define the feature values which ar over three standard deviations away
from the mean feature value. We also normalize all features to
zero-mean and unit standard deviation. We split the dataset into
\(N_{tr} = 15639\) training and \(N_{test} = 3910\) test examples
(80/20 split) and we fit a Neural Network with 3 hidden layers of 256,
128 and 36 units respectively. After 15 epochs using the Adam
optimizer with learning rate \(\eta = 0.02\), the model achieves a
normalized mean squared error (R-Squared) of \(0.25\) (0.75), which
corresponds to a MAE of \(37 \cdot 10^3\) dollars.

@@

\todo{Add Conclusion, limitations, and future work}
